Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, ``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data' and Machine Learning'', \emph{JCR}

### 2020-11-12 paper.tex

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, ``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data' and Machine Learning'', JCR},
            pdfauthor={Richard K. Morgan and Andreas Beger and Michael D. Ward},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020,
``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data'
and Machine Learning'', \emph{JCR}}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Richard K. Morgan and Andreas Beger and Michael D. Ward}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{03 June 2020}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\maketitle

\hypertarget{review-of-blair-and-sambanis-2020}{%
\section{Review of Blair and Sambanis
2020}\label{review-of-blair-and-sambanis-2020}}

Blair and Sambanis (2020) (BS for the sake of brevity hereafter) aim to
examine whether theory adds value to a forecasting model when compared
to non-parametric machine learning models, and specifically random
forests, whose specifications are not theory-informed. They find that it
does. They arrive at this conclusion by examining the problem of
predicting civil war onset, and find that a parsimonious model using a
small number of covariates derived from escalation theories of conflict
can forecast civil war onset better than alternative specifications
based on generic covariates not specifically informed by theory and a
kitchen sink model with more than 1,000 covariates.

BS specifically examine three questions:

\begin{itemize}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \tightlist
  \item
    how does the theoretically-driven Escalation model compare in
    forecast performance to alternative models not informed specifically
    by civil war onset theories,
  \end{enumerate}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \setcounter{enumi}{1}
  \tightlist
  \item
    does annual, structural information from the PITF instability
    forecasting model add to the Escalation model's monthly and 6-month
    predictions,
  \end{enumerate}
\item
  and (3) how accurate were predictions for the first half of 2016
  derived from the Escalation model?
\end{itemize}

To assess the first two questions BS use global datasets covering all
major countries from 2001 to 2015. Two versions of the dataset are used,
one at the country-month level, the other aggregated to 6-month
half-years. The main outcome is civil war onset, measured using
Sambanis' civil war dataset.

Both the first and second questions above rely on comparing the
Escalation model to various alternative models. The same procedure is
used in both cases:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Split the training data into training (2001 - 2007) and test (2008 -
  2015) sets.
\item
  Estimate the Escalation and other competing models.
\item
  Create out-of-sample (OOS) predictions from each model using the test
  set.
\item
  Calculate AUC-ROC measures for each set of OOS predictions.
\end{enumerate}

The corresponding results for each question are shown in BS Tables 1 and
2, which we will replicate further below.

To examine the first question, BS compare the test set of the Escalation
model to four alternative models. The independent variables for the
first set of analysis reported in Table 1 in the paper are all derived
from the ICEWS event data, using domestic events between actors within a
country. The models are:

\begin{itemize}
\tightlist
\item
  Escalation: a set of 10 theoretically-informed indicators.
\item
  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation,
  verbal conflict, verbal cooperation.
\item
  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from
  the ICEWS data for interactions between the government one one side
  and opposition or rebel actors on the other. These are directed, thus
  making for 4 total covariates.
\item
  CAMEO: counts for all CAMEO event codes, thus a total of 1,159
  covariates.
\item
  Average: unweighted average of the predictions from the 4 models
  above.
\end{itemize}

The results in Table 1, aside from the core base specification results,
include 8 additional robustness tests for both the 1-month and 6-month
versions. These robustness checks vary either (1) random forecast
hyperparameter values, or (2) the year used to split the train/test
data, or (3) alternative codings of the civil war onset dependent
variable.

The second question, whether structural variables add to the Escalation
model, is assessed by comparing the original escalation model to four
alternatives that incorporate annual, structural variables that are used
in the PITF instability forecasting model:

\begin{itemize}
\tightlist
\item
  Escalation Only: the original basic escalation model with only ICEWS
  predictors
\item
  With PITF Predictors: a random forest that as predictors has the
  escalation model indicators but also the PITF annual, structural
  variables
\item
  Weighted by PITF: escalation model predictions weighted using the PITF
  instability model predictions
\item
  PITF Split Population: the training data are split into high and low
  risk portions based on the PITF instability model predictions, two
  separate escalation random forests are trained on the splits, then
  re-combined into a single random forest that is used to create the
  test set predictions
\item
  PITF Only: a random forest model based only on the annual, structural
  PITF model predictors
\end{itemize}

The corresponding results are shown in BS Table 2.

Finally, BS used the Escalation model to create forcasts for the first
half of 2016, and in their third and final analysis, they score the
forecasts accuracy using civil war onset data later observed. This is
summarized in BS Table 3.

\hypertarget{replication-problems}{%
\section{Replication problems}\label{replication-problems}}

While attempting to replicate BS's results, we found several issues.

\hypertarget{smoothed-roc-curves}{%
\subsection{Smoothed ROC curves}\label{smoothed-roc-curves}}

The most consequential issue that we found is that all AUC-ROC values
reported in BS Tables 1 and 2 are calculated using smoothed ROC curves,
not the original, actual ROC curves. A reference to smoothing is made in
a single sentence in the paper (p.~12):

\begin{quote}
Figure 1 displays the corresponding ROC curves, smoothed for ease of
interpretation.
\end{quote}

This implies that the ROC curves were only smoothed in the references
Figure 1, but actually all AUC-ROC calculations throughout the
replication code use an option to smooth the ROC curves prior to AUC
calculation.

Figure 1 shows our replication of both the smoothed ROC curves BS
report, and the actual ROC curves on the right.

\begin{figure}
\centering
\includegraphics{figures/figure1-replicated.png}
\caption{Replication of BS Figure 1 with both smooth and non-smoothed
ROC curves.}
\end{figure}

ROC curves typically appear step-like in response to the distribution of
positive and negative cases in the data. In this case there are also
groups of cases with identical predicted probabilities, which accounts
for the unusual diagonal lines. In any case, with a sparse outcome like
civil war onsets, the true positive rate on the \emph{y}-axis only
changes when the prediction for a observed positive case is reached. For
these ROC curves, and for that matter in the basic train/test split used
for 12 of the 18 rows/models in BS Table 1, there are only 11 civil war
onset cases in the test set. Thus the ROC curves here are very
step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct
\emph{y} coordinates. We can only speculate, but maybe the very
``step''-like nature of the ROC curves plus the presence of unusual
diagonal lines accounts for the decision to use smoothed ROC curves.

As it turns out, this decision has a dramatic impact on the AUC-ROC
calculations. As we will show in the results further below, the
difference between the smoothed and original ROC AUC values is up to
0.12--a huge difference given that row-wise, the models being compared
typically differ by only 0.05 or less--and differentially impacts the
models that are being compared. In fact, BS's original results and
interpretation are entirely conditional on the use of smoothed AUC-ROC.

To our knowledge, the norm is to calculate AUC-ROC values using
original, not smoothed ROC curves. We are in fact not aware of other
work that uses smoothed ROC curves for AUC calculations. It might be
that there are theoretical reasons justifying the use of smoothed ROC
curves over the original ROC curves; but given that this decision
dramatically impacts the interpretation of the BS results, it minimally
would have warranted an explicit discussion in the paper. This is not
the case, and it is only mentioned in the sentence we quote above.

The next three issues we encountered all concern information in Table 2.

\hypertarget{incorrect-weighted-by-pitf-implementation}{%
\subsection{Incorrect ``Weighted by PITF''
implementation}\label{incorrect-weighted-by-pitf-implementation}}

The ``Weighted by PITF'' model is described as follows in BS, page 19:

@@

We infer that the intent is that the Escalation model's predictions for
the test set are weighted by the PITF model predictions for the test
set. The implementation actually weighs the test set predictions using
the PITF model predictions for the training set.\footnote{See
  \texttt{1mo\_run\_escalation\_weighted\_PITF.R} line 4, where the PITF
  predictions are taken from the training data set
  (\texttt{train\$pred\_prob\_plus1}). The next line is a hack extending
  the shorter \texttt{weight} vector with missing values to avoid a R
  warning when it is multiplied with the longer vector of escalation
  model test set predictions. Similarly in the 6 month version of this
  file.} This appears to be a coding error.

\hypertarget{incorrect-pitf-split-population-implementation}{%
\subsection{Incorrect ``PITF Split Population''
implementation}\label{incorrect-pitf-split-population-implementation}}

Similarly, the ``PITF Split Population'' model appears to be incorrectly
implemented. BS describe it as, on page 20:

\begin{quote}
The final approach is a random forest analog to split-population
modeling. first compute the average PITF predicted probability for each
country across all years in our training set. We define those that fall
in the bottom quartile as ``low risk'' and the rest as ``high risk.'' We
then run our escalation model on the high-risk and low-risk subsets
separately, combining the results into a single random forest (column
4).
\end{quote}

The intention clearly was to run two separate random forest models, one
each on the low- and high-risk training data splits. The replication
code does indeed run two separate random forecasts, but they are both
run on the exact same training data, which consists of the full training
data all other models are run on. The models are also identical
otherwise, i.e.~they use the same \emph{x} variables and the same random
forest hyper-parameter settings. The \emph{only} difference in the
models as they are implemented in the BS replication code is due to the
non-deterministic nature of the random forest model itself. If we ran
both with the same RNG seed, they would be identical in every respect,
producing identical predictions.\footnote{Disentangling this coding
  error is not straightforward as it occurs over several R scripts and
  requires (or at least is easier to verify by) running partway through
  the actual replication until the objects holding the training data for
  the models are instantiated and can be examined. We have documented
  details at
  \url{https://github.com/rickmorgan2/Blair-Sambanis-replication/issues/5}.}

The implementation error aside, this split-population analog model is
actually quite odd and does not actually replicate an analogue of the
idea behind split-population modeling. Although the two RFs are trained
on separate data (in our updated, fixed replication), the process of
combining them actually just creates a new, larger RF using both
component model's underlying decision (regression\footnote{See further
  below. Although the RF models are used for a binary decision problem,
  the actual implementation uses regression RFs for continuous outcomes.})
trees. Thus while all RF models throughout (except for one of the
robustness checks) are trained with 100,000 decision trees
(\texttt{ntree}), the new RF model after combination does indeed have
200,000 decision trees. Furthermore, the PITF model predictions do not
impact the way the combined RF model predicts at all, not even through a
binary low-/high-risk split. The split-population PITF RF model is
practically speaking just another Escalation model trained with
N=200,000 instead of N=100,000 trees and an extra odd randomization step
added to the already existing RF randomization facilities (row and
column sampling for each decision tree).

@@

Lastly, the AUC-ROC values reported in the original BS Table 2 are
calculated on the basis of slightly different numbers of underlying test
set cases (see Table \ref{table2-N}). ROC calculations for a set of
predictions can only be done on the set of cases for which both
non-missing predictions and non-missing outcomes are available. Those
sets differ across models (columns) for each row in Table 2. Thus a
difference in AUC-ROC values for two models could be due to the fact
that they were calculated on different sets of underlying cases, not
because the models are systemically performing at a different level. In
other words, the results for different models in BS Table 2 are actually
not comparable to the other, and any conclusions drawn from such
comparison are potentially incorrect.

We fix this issue by only using predictions for common joint subset of
cases that all models have non-missing predictions for. The original BS
Table 2 1-month have \emph{N}=11,806--12,495 versus a common joint
subset of 9,811, and for the 6-month row \emph{N}=2,070--2,233 versus
\emph{N}=1,915 for the common joint subset.

@@

BS show a confusiong matrix to score their 2016-H1 forecasts in Table 4.
Although the forecasts are for the probability of civil war onset, in
the replication code they are actually scored using the much more common
incidence of civil war, i.e.~including ongoing civil wars as ``1'''s.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict},
  pdfauthor={Andreas Beger; Richard K. Morgan; Michael D. Ward},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{floatrow}
\floatsetup[figure]{capposition=bottom}
\usepackage[format=plain,labelfont=bf,up]{caption}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict\thanks{John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at \url{https://github.com/andybega/Blair-Sambanis-replication}.}}
\author{Andreas Beger\footnote{Predictive Heuristics, \href{mailto:adbeger@gmail.com}{\nolinkurl{adbeger@gmail.com}}.} \and Richard K. Morgan\footnote{Varieties of Democracy Institute, University of Gothenburg, \href{mailto:rick.morgan2@gmail.com}{\nolinkurl{rick.morgan2@gmail.com}}.} \and Michael D. Ward\footnote{Predictive Heuristics, Duke University, University of Washington, \href{mailto:michael.don.ward@gmail.com}{\nolinkurl{michael.don.ward@gmail.com}}. Corresponding author.}}
\date{24 September 2020}

\begin{document}
\maketitle
\begin{abstract}
We examine the research protocols in Blair and Sambanis (nd), who argue that their theory-based model can predict civil war onsets better than several atheoretical alternatives or a model with country-structural factors. We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models. Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine. More importantly, we argue that rather than trying to contrast the roles of theory and "atheoretical" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant theories.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Blair and Sambanis (n.d., hereafter B\&S) argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed, they assert that with such theory forecasting is more accurate than has previously been possible. We re-examine the empirical basis for the claims made in support of it. We find that these claims to be unsupported and evidence presented for them to be incorrect. Their theory-based escalation model does not do better than the alternatives that they examine. It does worse. The reason for this is that they have made several mistakes in their research procedure. Further, the performance results they report are based on smoothed performance curves, not the empirical, unsmoothed curves. This provides misleading results. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 was incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars ``remarkably accurately''. Indeed, B\&S argue that this theoretical model outperforms four other ``more mechanical'' alternatives. They also claim that the integration of structure with process is better than alternatives over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{PITF (e.g., Goldstone et al. 2010; Beger and Ward 2017), the W-ICEWS project (Ward et al. 2013), VIEWS (Hegre et al. 2019), and others, e.g., Beger, Morgan, and Maxwell (2020).} B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

Before proceeding, we quote B\&S (page 24):

\begin{quote}
Our theoretically driven model generates accurate forecasts, with base specification AUCs of 0.82 and 0.85 over one- and six-month windows, respectively, and AUCs as high as 0.92 in other specifications. Our model also consistently and sometimes dramatically outperforms the alternatives we test. {[}\ldots{]} Cederman and Weidmann (2017, 476) argue that ``the hope that big data will somehow yield valid forecasts through theory-free `brute force' is misplaced in the area of political violence.'' Our results lend some credence to this claim.
\end{quote}

Although their analysis in the end does not support this point, we do not disagree that theory is important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with ``theoretical'' modeling, B\&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the space of extant theories for a phenomenon like civil war is so broad and non-specific in terms of an exact specification of what variables and in which functional form should constitute a prediction model, that it's both not very clear what the ``correct'' model should be, nor easy to demonstrate that any particular ``theoretical'' model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to ``explain'' a phenomenon to strengthen and whittle down the broad space of extant theories.

\hypertarget{summary-of-blair-and-sambanis-2020}{%
\section{Summary of Blair and Sambanis (2020)}\label{summary-of-blair-and-sambanis-2020}}

B\&S' analysis is based on the use of non-linear, non-deterministic machine learning models, and specifically random forests, one of which has a specification they argue is theory-based. Several others with more generic sets of covariates are also developed and compared. Notably, the analysis is at the country-month or country-(6 month) levels and relies in large part on indicators derived from the ICEWS event data. In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables. They arrive at this conclusion by examining the problem of predicting civil war onset. They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory, including a \textit{kitchen sink} model with more than 1,000 covariates.

B\&S specifically examine three questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  How does the theoretically-driven escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
\item
  Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
\item
  How accurate were predictions using the escalation model for the first half of 2016?
\end{enumerate}

To assess the first two questions, B\&S use ICEWS data covering all major countries from 2001 to 2015. Two versions of the dataset are used, one at the country-month level, the other aggregated to 6-month half-years. The main outcome variable is civil war onset, measured using Sambanis' civil war dataset.

Both the first and second questions above rely on comparing their escalation model to various alternative models. The same procedure is used in both cases:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Split the training data into training (2001 - 2007) and test (2008 - 2015) data.
\item
  Estimate the escalation and other competing models with the training data.
\item
  Create out-of-sample (OOS) predictions from each model using the test set.
\item
  Calculate AUC-ROC measures for each set of OOS predictions.
\end{enumerate}

To examine the first question, B\&S compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

\begin{itemize}
\tightlist
\item
  Escalation: a set of ten indicators, drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other.
\item
  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels. These are directed, thus making for four directed dyads, which with four quad categories make sixteen covariates.
\item
  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a total of four covariates.
\item
  CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling \(1,160\) covariates, which are mostly zero for any country in any month.
\item
  Average: unweighted average of the predictions from the four models briefly described above.
\end{itemize}

The corresponding results for each question are shown in B\&S Tables 1 and 2, which we examine further below. We accurately replicate their Tables 1 and 2. The results in Table 1, aside from the core base specification results, include eight additional robustness tests for both the 1-month and 6-month versions. These robustness checks vary either (1) random forecast hyperparameter values or (2) the year used to split the train/test data, or (3) alternative codings of the civil war onset dependent variable.

The second question, whether structural variables add to the escalation model, is assessed by comparing the original escalation model to four alternatives that incorporate annual, structural variables that are used in the PITF instability forecasting model:

\begin{itemize}
\tightlist
\item
  Escalation Only: the original basic escalation model with only ICEWS predictors.
\item
  With PITF Predictors: a random forest that also adds the PITF annual, structural variables.
\item
  Weighted by PITF: escalation model predictions weighted using the PITF instability model predictions.
\item
  PITF Split Population: the training data are split into high and low risk portions based on the PITF instability model predictions, two separate escalation random forests are trained on the splits, then re-combined into a single random forest that is used to create the test set predictions.
\item
  PITF Only: a random forest model based only on the annual, structural PITF model predictors.
\end{itemize}

The corresponding results are shown in B\&S Table 2.

Finally, B\&S used their escalation model to create forecasts for the first half of 2016. In their third analysis, they score the forecasts accuracy using subsequent civil war onset data. This is summarized in B\&S Table 3.

\hypertarget{implementation-issues-in-bs}{%
\section{Implementation Issues in B\&S}\label{implementation-issues-in-bs}}

While replicating and analyzing B\&S's results, we found several issues worthy of further discussion and investigation. These are (a) the use of smoothed ROC curves to draw conclusions about which model is best, (b) incorrect implementations of the weighted by PITF and PITF analog split-population models, (c) inconsistent test sets for the models examined, and (d) incorrect scoring of the 2016 forecasts.\footnote{We also note there are additional concerns arising from the question of how the random forest models were tuned by B\&S, especially given the way they are used is unorthodox. We did not further investigate the latter issue as it is rendered somewhat moot by the changes in results after addressing the preceding issues.}

We believe that these research decisions and issues lead B\&S to incorrect conclusions. The escalation model \textit{is not} the best; it actually performs worse than the atheoretical, kitchen sink model with over \(1000\) variables. We discuss these five issues below. We defer a complete analysis that corrects all these issues until later, as there are many possible permutations of a seriatim unfolding.

\hypertarget{smoothed-roc-curves}{%
\subsection{Smoothed ROC Curves}\label{smoothed-roc-curves}}

The most consequential issue that we found is that all AUC-ROC values reported in B\&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. B\&S refer to smoothing only in the context of the ROC curves in their Figure 1, justified as easing interpretation. However, smoothing is also used in all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs. We examined all 63 references in B\&S and also 15 articles in two randomly picked issues of JCR for 2020 (numbers 1 and 9), in order to survey the use of ROC curves. Of the 37 articles that used a binary outcome, 19 included either ROC curve figures or a table with AUC-ROC values. We found that 90\%--17 of 19--clearly used empirical ROC curves, and 2 (10\%) where we could not establish whether empirical or smoothed curves were used.\footnote{One paper reports AUC-ROC values but no figure with ROC curves, and we could not find publicly posted replication code. The other is an ambiguous case where ROC curves are manually constructed in a way that might be considered "smoothing"--and one of the two figures certainly looks smoothed--but it's not a parametric smoothing method like used in B\&S. The online replication archive has more details at https://github.com/andybega/Blair-Sambanis-replication/tree/master/journal-survey. Both of the ambiguous cases include Blair as coauthor.} The survey results are not surprising since the original ideal as well as contemporaneous documentation suggest this is the gold standard. What is surprising is that anyone would choose the smoothed ROC implementation.

The next three issues we encountered all concern information in B\&S Table 2, which is the key table in which B\&S report AUC-ROC values to demonstrate the ``Escalation'' models superiority.

\hypertarget{incorrect-weighted-by-pitf-implementation}{%
\subsection{Incorrect ``Weighted by PITF'' Implementation}\label{incorrect-weighted-by-pitf-implementation}}

The ``Weighted by PITF'' model is described as follows in B\&S, page 19:

@@

We believe that B\&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.

As a result: (1) in the test set, all countries with names that alphabetically start after ``Tongo'' are dropped from the data, including Turkey, Ukraine, Yemen, etc.; and (2) the cases whose PITF training data predictions are combined with Escalation model test predictions in the weighting are nonsensical. For example, the ``Weighted by PITF'' model predictions for Syria from 2008 to 2015 are the Escalation test predictions for that time period weighted, respectively, by the 2002 to 2007 PITF predictions for South Africa and the 2001 to 2002 PITF predictions for Zambia.

This appears to be an easily corrected coding error on the part of B\&S.

\hypertarget{incorrect-pitf-split-population-implementation}{%
\subsection{Incorrect ``PITF Split Population'' Implementation}\label{incorrect-pitf-split-population-implementation}}

The ``PITF Split Population'' model also appears to be incorrectly implemented, owing to a coding error. B\&S describe it on page 20:

\begin{quote}
The final approach is a random forest analog to split-population modeling. We first compute the average PITF predicted probability for each country across all years in our training set. We define those that fall in the bottom quartile as ``low risk'' and the rest as ``high risk.'' We then run our escalation model on the high-risk and low-risk subsets separately, combining the results into a single random forest {[}\ldots{]}.
\end{quote}

This description suggests that B\&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the \emph{exact same training data}, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.

The model specifications are also identical; i.e., they use the same \emph{x} variables and the same random forest hyper-parameter settings. The \emph{only} difference in the models as they are implemented in the B\&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing two forests of identical decision trees and, thus, identical predictions.

The implementation error aside, this split-population analog model is quite odd and does not replicate the idea behind split-population modeling (Chiba, Metternich, and Ward 2015; Beger et al. 2017). Although the two random forest models are trained on separate data (in our updated, fixed replication), the process of combining them actually creates a new, larger random forest using both component model's underlying decision trees. Thus, while all random forest models throughout (except for one of the robustness checks) are trained with 100,000 decision trees (\texttt{ntree}), the new random forest model after combination does have 200,000 decision trees. Furthermore, the PITF model predictions do not impact how the combined random forest model predicts at all, not even through a binary low-/high-risk split. The split-population PITF random forest model is practically speaking simply another escalation model trained with N=200,000 instead of N=100,000 trees and an extra odd randomization step added to the already existing random forest randomization facilities (row and column sampling for each decision tree). This does not adequately implement their split-sample modeling strategy.

@@

Further, the AUC-ROC values reported in the original B\&S Table 2 are calculated on different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in B\&S Table 2.

Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B\&S Table 2 are not comparable to one another, and conclusions drawn from such comparison are potentially incorrect.

@@

B\&S present a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war \emph{onset}, in the replication code, they are actually scored using \emph{incidence} of civil war. By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict. When forecasting rare events, it is common to have many false positives, i.e.~instances where we predict onset but no onsets occur. \emph{Ex ante}, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).

\hypertarget{results-of-the-updated-analysis}{%
\section{Results of the updated analysis}\label{results-of-the-updated-analysis}}
